---
title: '**Evaluating the Suitability of Lithic Illustrations in Morphometric Analyses**'
author: Christian Steven Hoggard, Thomas Birch, Cory Marie Stade, Katrien Janin and
  Felix Riede
output:
  pdf_document: default
  word_document: default
  html_document: default
---
<style>
body {
text-align: justify}
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
  
### **Abstract**  

Illustrations of lithic artefacts are an abundant source of morphological and technological information for those interested in our human past. As a typical part of archaeological reports and publications, lithic drawings are - or have to be - trusted as faithful reproductions of the selected artefacts. Despite the considerable epistemic work lithic illustrations (and illustrators) are expected to do, usually little information is available regarding the illustrator's technical skill; thus, it remains unknown whether drawings produced by illustrators of differing technical skill are comparable or produce images of equal analytical potential to other media, e.g. photographs. The issue of lithic illustration accuracy is brought to the fore by the recent emergence of geometric morphometric approaches as innovative and powerful ways of describing and analysing complex shapes, as lithic illustrations provide one of the key sources for such analyses. Motivated by these issues, we present an experiment investigating the degree of error observed in illustrations of differing technical illustrative skill. Analyses suggest that lithic illustrations produced by individuals with a variety of experience in drawing lithics create, in the majority of instances, equally faithful representations (in outline shape) of chipped stone artefacts. With error observed in a small number of instances, archaeologists are still urged to be critical of an illustration's source prior to lineal and geometric morphometric methodologies. Despite this, archaeologists can be confident in their exactitude and we remain strong advocates in favour of lithic illustrations as a readily available legacy resource for morphometric analyses.
  
  
### **Introduction**  

This is a R Markdown documented associated with the article *Evaluating the Suitability of Lithic Illustrations in Morphometric Analyses*, original research for *The Journal of the Lithic Studies Society* (http://www.lithics.org/). This Markdown document details the analytical procedure used throughout the article, from data importing and the packages used, to the univariate and multivariate framework underpinning the article's findings. 

### **Data files**  

For this markdown, a number of different files are necessary:   
1. The three raw landmark files (for handaxe, tanged point and elongated artefacts), these are: **elongated.tps**, **handaxe.tps** and **tanged.tps**  
2. The three data frames for the respective landmark files: **elongated.csv**, **handaxe.csv** and **tanged.csv**  
3. The raw measurement length and width data for all artefacts: **measurement_data.csv**  
4) The files necessary to investigate landmark digitisation error: **digitisation_error.tps** and **digitisation_error.csv**  
5) The file necessary to investigate measurement error: **measurement_error.csv**  
6) The table of links for all landmark configurations:  **curveslide.csv**  

A copy of all files can also be found on GitHub and the Open Science Framework (OSF).  
GitHub: https://github.com/CSHoggard/-Lithic_Illustrations  
OSF: https://osf.io/xtghn/  

### **Stage 1: Package installation and data extraction**  

For this article, ten packages are required:  
1. **psych v.1.8.12**  https://cran.r-project.org/web/packages/psych/index.html  
2. **geomorph v.3.1.2** https://cran.r-project.org/web/packages/geomorph/index.html  
3. **tidyverse v.1.2.1** https://cran.r-project.org/web/packages/tidyverse/index.html  
4. **vegan v.2.5-4** https://cran.r-project.org/web/packages/vegan/index.html    
5. **MASS v.7.3-51.4** https://cran.r-project.org/web/packages/MASS/index.html   
6. **cowplot v.0.9.3** https://cran.r-project.org/web/packages/cowplot/index.html  
7. **ggpubr v.0.9.3**  https://cran.r-project.org/web/packages/ggpubr/index.html  
8. **rio v.0.5.16**  https://cran.r-project.org/web/packages/rio/index.html  
9. **LaMBDA v.0.1.0.9000** https://rdrr.io/github/akiopteryx/lambda/  
10. **devtools v.2.3.1** https://cran.r-project.org/web/packages/devtools/index.html  

The latest versions of each package can be installed and sourced through the following code:    

```{r, echo = TRUE, eval = FALSE}
if(!require("psych")) install.packages('psych', repos='http://cran.us.r-project.org')
if(!require("geomorph")) install.packages('geomorph', repos='http://cran.us.r-project.org')
if(!require("tidyverse")) install.packages('tidyverse', repos='http://cran.us.r-project.org')
if(!require("vegan")) install.packages('vegan', repos='http://cran.us.r-project.org')
if(!require("MASS")) install.packages('MASS', repos='http://cran.us.r-project.org')
if(!require("cowplot")) install.packages('cowplot', repos='http://cran.us.r-project.org')
if(!require("ggpubr")) install.packages('ggpubr', repos='http://cran.us.r-project.org')
if(!require("rio")) install.packages('rio', repos='http://cran.us.r-project.org')
```

The `LaMBDA` package (LandMark-Based Data Assessment) is not on CRAN and requires downloading from GitHub (via the `devtools` package):    

```{r, echo = TRUE, eval = FALSE}
install.packages(“devtools”)
devtools::install_github(“akiopteryx/lambda”)
```

Should specific functions become outdated then please refer to the package versions as cited above. Once downloaded, activate all packages:

```{r, echo = TRUE, eval = TRUE, warning = FALSE, message = FALSE}
library(psych) ### load the listed package
library(geomorph) ### load the listed package
library(tidyverse) ### load the listed package
library(vegan) ### load the listed package
library(MASS) ### load the listed package
library(cowplot) ### load the listed package
library(ggpubr) ### load the listed package
library(LaMBDA) ### load the listed package
library(rio) ### load the listed package
```

### **Stage 1: Bringing all files into the R Environment**  

Given the large number of files required for the individual analyses (shape analysis, metric analysis, digitisation error and measurement error), and to prevent issues of working directories, all files are brought into the R Environment as .rds objects; these are objects which were imported into R and stored on GitHub (and downloaded when required). Using the `rio::import()` function allows all objects to be brought into the workspace.  

```{r, echo = TRUE, eval = TRUE}
landmarks_elongated <- import("https://github.com/CSHoggard/-Lithic_Illustrations/raw/master/landmarks_elongated.rds")
landmarks_handaxe <- import("https://github.com/CSHoggard/-Lithic_Illustrations/raw/master/landmarks_handaxe.rds")
landmarks_tanged <- import("https://github.com/CSHoggard/-Lithic_Illustrations/raw/master/landmarks_tanged.rds")
shape_data_elongated <- import("https://github.com/CSHoggard/-Lithic_Illustrations/raw/master/shape_data_elongated.rds")
shape_data_tanged <- import("https://github.com/CSHoggard/-Lithic_Illustrations/raw/master/shape_data_tanged.rds")
shape_data_handaxe <- import("https://github.com/CSHoggard/-Lithic_Illustrations/raw/master/shape_data_handaxe.rds")
metric_data <- import("https://github.com/CSHoggard/-Lithic_Illustrations/raw/master/metric_data.rds")
digitisation_error_landmarks <- import("https://github.com/CSHoggard/-Lithic_Illustrations/raw/master/digitisation_error_landmarks.rds")
digitisation_error_landmarks_data <- import("https://github.com/CSHoggard/-Lithic_Illustrations/raw/master/digitisation_error_landmarks_data.rds")
digitisation_error_metrics <- import("https://github.com/CSHoggard/-Lithic_Illustrations/raw/master/digitisation_error_metrics.rds")
shape_data_sliders <- import("https://github.com/CSHoggard/-Lithic_Illustrations/raw/master/shape_data_sliders.rds")
```  
  
### **Stage 2: Sources of Error and Investigating Landmark Count**  

### **Stage 2A: Digitisation and Landmark Error**  

A Procrustes ANOVA (see in-text reference) was performed to calculate error associated with the digitisation of landmarks.   

```{r, echo = TRUE, eval = TRUE, message = FALSE}
gpa_digi_error <- gpagen(digitisation_error_landmarks, Proj = TRUE, curves =  shape_data_sliders, ProcD = TRUE, surfaces = NULL, print.progress = FALSE) ### formulates a generalised Procrustes analysis (GPA) file for repeated specimens
gpa_digi_error  ### calls the GPA landmark configuration 
plot(gpa_digi_error) ### visualisation of the landmark configuration

gpa_digi_error_df <- geomorph.data.frame(gpa_digi_error, attempt = digitisation_error_landmarks_data$Attempt)
gpaprocD <- procD.lm(coords ~ attempt, data = gpa_digi_error_df, print.progress = FALSE) ### ANOVA (coordinates vs. attempt)
summary(gpaprocD) ### summary
gpaprocD$aov.table$SS[1]/gpaprocD$aov.table$SS[3]*100 ### error expressed as a percentage (8.599812%)
```
  
### **Stage 2B: Measurement Error**

To calculate error associated with lineal measurements fractional uncertainity (standard error divided by the mean) was calculated:  

```{r, echo = TRUE, eval = TRUE}
head(digitisation_error_metrics) 
statsl  <- describe(digitisation_error_metrics$Length_mm) ### descriptive statistics
statsw  <- describe(digitisation_error_metrics$Width_mm) ### descriptive statistics
statssf <- describe(digitisation_error_metrics$Scale_Factor) ### descriptive statistics

(statsl$se/statsl$mean)   * 100 ### fractional uncertainity (length)
(statsw$se/statsw$mean)   * 100 ### fractional uncertainity (width)
(statssf$se/statssf$mean) * 100 ### fractional uncertainity (scale factor: as calibrated through tpsDig2)
```

### **Stage 2C: Investigating Landmark Number**  

The `LaMBDA::LaSEC()` function is used to assess the fidelity of morphological characterisation. This function helps the user to identify under- and over-sampling of landmarks, the robustness of the characterisation and determine how many landmarks can be removed without compromising the necessary shape information. A two-dimensional array of each landmark file was used, with 500 iterations, to determine if twenty landmarks were suitable. 500 iterations were used for the analysis, here 10 iterations are exemplified (for ease and speed):

```{r, echo = TRUE, eval = TRUE}
lasec(two.d.array(landmarks_elongated), 2, iter = 10, show.progress = F) ### may take some time
lasec(two.d.array(landmarks_tanged), 2, iter = 10, show.progress = F) ### may take some time
lasec(two.d.array(landmarks_handaxe), 2, iter = 10, show.progress = F) ### may take some time
```
 
### **Stage 3: Generalised Procrustes Analysis (GPA)**

With all errors calculated and deemed insignificant, we can now transform each artefact class through a GPA using the `geomorph::gpagen()` function as above:

```{r, echo = TRUE, eval = TRUE}
gpa_elongated <- gpagen(landmarks_elongated, Proj = TRUE, ProcD = TRUE, curves = shape_data_sliders, surfaces = NULL, print.progress = FALSE) ### gpa (using a procrustes distance criterion): all elongated artefacts
plot(gpa_elongated) ### plots the procrustes coordinates for all elongated artefacts

gpa_tanged <- gpagen(landmarks_tanged, Proj = TRUE, ProcD = TRUE, curves = shape_data_sliders, surfaces = NULL, print.progress = FALSE) ### gpa (procrustes distance criterion): all tanged artefacts
plot(gpa_tanged) ### plots the procrustes coordinates for all tanged artefacts

gpa_handaxe <- gpagen(landmarks_handaxe, Proj = TRUE, ProcD = TRUE, curves = shape_data_sliders, surfaces = NULL, print.progress = FALSE) ### gpa (Procrustes distance criterion): all handaxe artefacts
plot(gpa_handaxe) ### plots the procrustes coordinates for all handaxe artefacts
```

### **Stage 4: Exploratory Analysis through Principal Component Analysis (PCA)**

Individual PCAs for all artefact classes were created to visualise differences in illustration skill between and within the different artefacts. PCAs were made in geomorph, through the `geomorph:gm.prcomp()` function, before being exported and modified (for better visualisation) in the tidyverse Note: specific shape differences can be calculated through the created files. The visualisations can then be exported through the `tidyverse:ggsave()` function. 

```{r, echo = TRUE, eval = TRUE}
pca_elongated <- gm.prcomp(gpa_elongated$coords) ### pca (geomorph)
pca_elongated ### pca summary
elongated_ds <- cbind(shape_data_elongated, pca_elongated$x) ### tidyverse compatible format

pca_tanged <- gm.prcomp(gpa_tanged$coords) ### pca (geomorph)
pca_tanged ### pca summary
tanged_ds <- cbind(shape_data_tanged, pca_tanged$x) ### tidyverse compatible format

pca_handaxe <- gm.prcomp(gpa_handaxe$coords) ### pca (geomorph)
pca_handaxe ### pca summary
handaxe_ds <- cbind(shape_data_handaxe, pca_handaxe$x) ### tidyverse compatible format

ggplot(elongated_ds) + 
  geom_point(aes(x = Comp1, y = Comp2, colour = Artefact, shape = Class), size = 2) + 
  labs(x = "Principal Component 1 (45.816%)", y = "Principal Component 2 (24.298%)", shape = "Method/skill", colour = "Artefact") + 
  scale_color_manual(values=c("#e41a1c","#377eb8","#4daf4a","#984ea3","#ff7f00")) +  
  scale_shape_manual(values=c(17,15,3,16)) +
  theme_minimal() ### figure 5 creation

ggplot(tanged_ds) + 
  geom_point(aes(x = Comp1, y = Comp2, colour = Artefact, shape = Class), size = 2) + 
  labs(x = "Principal Component 1 (50.805%)", y = "Principal Component 2 (33.926%)", shape = "Method/skill", colour = "Artefact") + 
  scale_color_manual(values=c("#e41a1c","#377eb8","#4daf4a","#984ea3","#ff7f00")) +  
  scale_shape_manual(values=c(17,15,3,16)) + 
  theme_minimal() ### figure 6 creation

ggplot(handaxe_ds) + 
  geom_point(aes(x = Comp1, y = Comp2, colour = Artefact, shape = Class), size = 2) + 
  labs(x = "Principal Component 1 (84.840%)", y = "Principal Component 2 (6.987%)", shape = "Method/skill", colour = "Artefact") + 
  scale_color_manual(values=c("#e41a1c","#377eb8","#4daf4a","#984ea3","#ff7f00")) +  
  scale_shape_manual(values=c(17,15,3,16)) + 
  theme_minimal() ### figure 7 creation

```

### **Stage 5: Further analysis (MANOVA and Discriminant Analysis)**

Each individual artefact (and the collections of drawings for each artefact) was then analysed to examine whether differences in illustrator skill were observed. This was first done through a Procrustes ANOVA, providing a statistical framework for detecting any such difference. Discriminant analyses were then conducted for each artefact to test for between-group variance. The code for each artefact follows this order:  

1) The Procrustes coordinates from an artefact are extracted from the dataframe.  
2) A geomorph-specific data-frame is created through the `geomorph::geomorph.data.frame()` function, with the Procrustes coordinates and the class factor.  
3) A Procrustes ANOVA is then performed through the `geomorph::procD.lm()` function, with the results summarised through the `base::summary()` argument.  
4) A new data frame is generated to produce the discriminant analysis (through the `MASS::lda() function`).  
5) The subsequent data is visualised through the `tidyverse` package.  
6) This is repeated for each artefact within each artefact class.  


```{r, echo = TRUE, eval = TRUE, warnings = FALSE, message = FALSE}
elongated1 <- gpa_elongated$coords[, , 1:18] 
df_elongated1 <- geomorph.data.frame(shape = elongated1, class = shape_data_elongated$Class[1:18], artefact = shape_data_elongated$Artefact[1:18]) 
E1 <- procD.lm(shape ~ class, data = df_elongated1, print.progress = FALSE) 
summary(E1) 

elongated1pcs <- as.data.frame(pca_elongated$x[1:18, 1:10])
elongated1class <- shape_data_elongated$Class[1:18] 
elongated1pcs <- cbind(elongated1pcs, elongated1class)
elongated1pcs <- rename(elongated1pcs, Class = elongated1class)
elongated1lda <- lda(Class ~ ., data = elongated1pcs)
elongated1ldapredict <- predict(elongated1lda) 
elongated1ldaplot <- cbind(elongated1ldapredict$x[, 1:3], elongated1pcs) 
elongated1ldaplotggplot <- ggplot(elongated1ldaplot, aes(LD1, LD2)) + 
  geom_point(aes(shape = Class), size = 2) + 
  labs(x = "LDA 1", y = "LDA 2") + 
  scale_shape_manual(values=c(17,15,3,16)) + 
  theme_minimal() + 
  theme(axis.text.x = element_blank(), 
        axis.text.y = element_blank(),
        axis.title.x = element_text(size = 6),
        axis.title.y = element_text(size = 6), 
        legend.position = "none")

elongated2 <- gpa_elongated$coords[, , 19:36]
df_elongated2 <- geomorph.data.frame(shape = elongated2, class = shape_data_elongated$Class[19:36], artefact = shape_data_elongated$Artefact[19:36])
E2 <- procD.lm(shape ~ class, data = df_elongated2, print.progress = FALSE)
summary(E2)
 
elongated2pcs <- as.data.frame(pca_elongated$x[19:36, 1:10])
elongated2class <- shape_data_elongated$Class[19:36]
elongated2pcs <- cbind(elongated2pcs, elongated2class)
elongated2pcs <- rename(elongated2pcs, Class = elongated2class)
elongated2lda <- lda(Class ~ ., elongated2pcs)
elongated2ldapredict <- predict(elongated2lda)
elongated2ldaplot <- cbind(elongated2ldapredict$x[, 1:3], elongated2pcs)
elongated2ldaplotggplot <- ggplot(elongated2ldaplot, aes(LD1, LD2)) + 
  geom_point(aes(shape = Class), size = 2) + 
  labs(x = "LDA 1", y = "LDA 2") + 
  scale_shape_manual(values=c(17,15,3,16)) + 
  theme_minimal() + 
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        legend.position = "none",
        axis.title.y = element_text(size = 6),
        axis.title.x = element_text(size = 6))

elongated3 <- gpa_elongated$coords[, , 37:54]
df_elongated3 <- geomorph.data.frame(shape = elongated3, class = shape_data_elongated$Class[37:54], artefact = shape_data_elongated$Artefact[37:54])
E3 <- procD.lm(shape ~ class, data = df_elongated3, print.progress = FALSE)
summary(E3) 

elongated3pcs <- as.data.frame(pca_elongated$x[37:54, 1:10])
elongated3class <- shape_data_elongated$Class[37:54]
elongated3pcs <- cbind(elongated3pcs, elongated3class)
elongated3pcs <- rename(elongated3pcs, Class = elongated3class)
elongated3lda <- lda(Class ~ ., elongated3pcs)
elongated3ldapredict <- predict(elongated3lda)
elongated3ldaplot <- cbind(elongated3ldapredict$x[, 1:3], elongated3pcs)
elongated3ldaplotggplot <- ggplot(elongated3ldaplot, aes(LD1, LD2)) +
  geom_point(aes(shape = Class), size = 2) +
  labs(x = "LDA 1", y = "LDA 2") +
  scale_shape_manual(values=c(17,15,3,16)) +
  theme_minimal() + 
  theme(axis.text.x = element_blank(), 
        axis.text.y = element_blank(), 
        legend.position = "none",  
        axis.title.y = element_text(size = 6),  
        axis.title.x = element_text(size = 6))

elongated4 <- gpa_elongated$coords[, , 55:72]
df_elongated4 <- geomorph.data.frame(shape = elongated4, class = shape_data_elongated$Class[55:72], artefact = shape_data_elongated$Artefact[55:72])
E4 <- procD.lm(shape ~ class, data = df_elongated4, print.progress = FALSE)
summary(E4)

elongated4pcs <- as.data.frame(pca_elongated$x[55:72, 1:10])
elongated4class <- shape_data_elongated$Class[55:72]
elongated4pcs <- cbind(elongated4pcs, elongated4class)
elongated4pcs <- rename(elongated4pcs, Class = elongated4class)
elongated4lda <- lda(Class ~ ., elongated4pcs)
elongated4ldapredict <- predict(elongated4lda)
elongated4ldaplot <- cbind(elongated4ldapredict$x[, 1:3], elongated4pcs)
elongated4ldaplotggplot <- ggplot(elongated4ldaplot, aes(LD1, LD2)) +
  geom_point(aes(shape = Class), size = 2) +
  labs(x = "LDA 1", y = "LDA 2") +
  scale_shape_manual(values=c(17,15,3,16)) +
  theme_minimal() + 
  theme(axis.text.x = element_blank(), 
        axis.text.y = element_blank(), 
        legend.position = "none",  
        axis.title.y = element_text(size = 6),  
        axis.title.x = element_text(size = 6))

elongated5 <- gpa_elongated$coords[, , 73:90] 
df_elongated5 <- geomorph.data.frame(shape = elongated5, class = shape_data_elongated$Class[73:90], artefact = shape_data_elongated$Artefact[73:90]) 
E5 <- procD.lm(shape ~ class, data = df_elongated5, print.progress = FALSE)
summary(E5)

elongated5pcs <- as.data.frame(pca_elongated$x[73:90, 1:10])
elongated5class <- shape_data_elongated$Class[73:90]
elongated5pcs <- cbind(elongated5pcs, elongated5class)
elongated5pcs <- rename(elongated5pcs, Class = elongated5class)
elongated5lda <- lda(Class ~ ., elongated5pcs)
elongated5ldapredict <- predict(elongated5lda)
elongated5ldaplot <- cbind(elongated5ldapredict$x[, 1:3], elongated5pcs)
elongated5ldaplotggplot <- ggplot(elongated5ldaplot, aes(LD1, LD2)) +
  geom_point(aes(shape = Class), size = 2) +
  labs(x = "LDA 1", y = "LDA 2") +
  scale_shape_manual(values=c(17,15,3,16)) +
  theme_minimal() + 
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        legend.position = "none",
        axis.title.y = element_text(size = 6),
        axis.title.x = element_text(size = 6))

tanged1 <- gpa_tanged$coords[, , 1:18]
df_tanged1 <- geomorph.data.frame(shape = tanged1, class = shape_data_tanged$Class[1:18], artefact = shape_data_tanged$Artefact[1:18])
T1 <- procD.lm(shape ~ class, data = df_tanged1, print.progress = FALSE)
summary(T1) 

tanged1pcs <- as.data.frame(pca_tanged$x[1:18, 1:10])
tanged1class <- shape_data_tanged$Class[1:18]
tanged1pcs <- cbind(tanged1pcs, tanged1class)
tanged1pcs <- rename(tanged1pcs, Class = tanged1class)
tanged1lda <- lda(Class ~ ., tanged1pcs)
tanged1ldapredict <- predict(tanged1lda)
tanged1ldaplot <- cbind(tanged1ldapredict$x[, 1:3], tanged1pcs)
tanged1ldaplotggplot <- ggplot(tanged1ldaplot, aes(LD1, LD2)) + 
  geom_point(aes(shape = Class), size = 2) +
  labs(x = "LDA 1", y = "LDA 2") +
  scale_shape_manual(values=c(17,15,3,16)) +
  theme_minimal() + 
  theme(axis.text.x = element_blank(), 
        axis.text.y = element_blank(), 
        legend.position = "none",  
        axis.title.y = element_text(size = 6),  
        axis.title.x = element_text(size = 6))

tanged2 <- gpa_tanged$coords[, , 19:36]
df_tanged2 <- geomorph.data.frame(shape = tanged2, class = shape_data_tanged$Class[19:36], artefact = shape_data_tanged$Artefact[19:36])
T2 <- procD.lm(shape ~ class, data = df_tanged2, print.progress = FALSE)
summary(T2)

tanged2pcs <- as.data.frame(pca_tanged$x[19:36, 1:10])
tanged2class <- shape_data_tanged$Class[19:36]
tanged2pcs <- cbind(tanged2pcs, tanged2class)
tanged2pcs <- rename(tanged2pcs, Class = tanged2class)
tanged2lda <- lda(Class ~ ., tanged2pcs)
tanged2ldapredict <- predict(tanged2lda)
tanged2ldaplot <- cbind(tanged2ldapredict$x[, 1:3], tanged2pcs)
tanged2ldaplotggplot <- ggplot(tanged2ldaplot, aes(LD1, LD2)) +
  geom_point(aes(shape = Class), size = 2) +
  labs(x = "LDA 1", y = "LDA 2") +
  scale_shape_manual(values=c(17,15,3,16)) +
  theme_minimal() + 
  theme(axis.text.x = element_blank(), 
        axis.text.y = element_blank(), 
        legend.position = "none",  
        axis.title.y = element_text(size = 6),  
        axis.title.x = element_text(size = 6))

tanged3 <- gpa_tanged$coords[, , 37:54] 
df_tanged3 <- geomorph.data.frame(shape = tanged3, class = shape_data_tanged$Class[37:54], artefact = shape_data_tanged$Artefact[37:54])
T3 <- procD.lm(shape ~ class, data = df_tanged3, print.progress = FALSE)
summary(T3)

tanged3pcs <- as.data.frame(pca_tanged$x[37:54, 1:10])
tanged3class <- shape_data_tanged$Class[37:54]
tanged3pcs <- cbind(tanged3pcs, tanged3class)
tanged3pcs <- rename(tanged3pcs, Class = tanged3class)
tanged3lda <- lda(Class ~ ., tanged3pcs)
tanged3ldapredict <- predict(tanged3lda)
tanged3ldaplot <- cbind(tanged3ldapredict$x[, 1:3], tanged3pcs)
tanged3ldaplotggplot <- ggplot(tanged3ldaplot, aes(LD1, LD2)) +
  geom_point(aes(shape = Class), size = 2) +
  labs(x = "LDA 1", y = "LDA 2") +
  scale_shape_manual(values=c(17,15,3,16)) +
  theme_minimal() + 
  theme(axis.text.x = element_blank(), 
        axis.text.y = element_blank(), 
        legend.position = "none",  
        axis.title.y = element_text(size = 6),  
        axis.title.x = element_text(size = 6))

tanged4 <- gpa_tanged$coords[, , 55:72]
df_tanged4 <- geomorph.data.frame(shape = tanged4, class = shape_data_tanged$Class[55:72], artefact = shape_data_tanged$Artefact[55:72])
T4 <- procD.lm(shape ~ class, data = df_tanged4, print.progress = FALSE)
summary(T4)

tanged4pcs <- as.data.frame(pca_tanged$x[55:72, 1:10])
tanged4class <- shape_data_tanged$Class[55:72]
tanged4pcs <- cbind(tanged4pcs, tanged4class)
tanged4pcs <- rename(tanged4pcs, Class = tanged4class)
tanged4lda <- lda(Class ~ ., tanged4pcs)
tanged4ldapredict <- predict(tanged4lda)
tanged4ldaplot <- cbind(tanged4ldapredict$x[, 1:3], tanged4pcs)
tanged4ldaplotggplot <- ggplot(tanged4ldaplot, aes(LD1, LD2)) +
  geom_point(aes(shape = Class), size = 2) +
  labs(x = "LDA 1", y = "LDA 2") +
  scale_shape_manual(values=c(17,15,3,16)) +
  theme_minimal() + 
  theme(axis.text.x = element_blank(), 
        axis.text.y = element_blank(), 
        legend.position = "none",  
        axis.title.y = element_text(size = 6),  
        axis.title.x = element_text(size = 6))

tanged5 <- gpa_tanged$coords[, , 73:90]
df_tanged5 <- geomorph.data.frame(shape = tanged5, class = shape_data_tanged$Class[73:90], artefact = shape_data_tanged$Artefact[73:90])
T5 <- procD.lm(shape ~ class, data = df_tanged5, print.progress = FALSE)
summary(T5)

tanged5pcs <- as.data.frame(pca_tanged$x[73:90, 1:10])
tanged5class <- shape_data_tanged$Class[73:90]
tanged5pcs <- cbind(tanged5pcs, tanged5class)
tanged5pcs <- rename(tanged5pcs, Class = tanged5class)
tanged5lda <- lda(Class ~ ., tanged5pcs)
tanged5ldapredict <- predict(tanged5lda)
tanged5ldaplot <- cbind(tanged5ldapredict$x[, 1:3], tanged5pcs)
tanged5ldaplotggplot <- ggplot(tanged5ldaplot, aes(LD1, LD2)) +
  geom_point(aes(shape = Class), size = 2) +
  labs(x = "LDA 1", y = "LDA 2") + 
  scale_shape_manual(values=c(17,15,3,16)) +
  theme_minimal() + 
  theme(axis.text.x = element_blank(), 
        axis.text.y = element_blank(), 
        legend.position = "none",  
        axis.title.y = element_text(size = 6),  
        axis.title.x = element_text(size = 6))

handaxe1 <- gpa_handaxe$coords[, , 1:18]
df_handaxe1 <- geomorph.data.frame(shape = handaxe1, class = shape_data_handaxe$Class[1:18], artefact = shape_data_handaxe$Artefact[1:18])
H1 <- procD.lm(shape ~ class, data = df_handaxe1, print.progress = FALSE)
summary(H1)

handaxe1pcs <- as.data.frame(pca_handaxe$x[1:18, 1:10])
handaxe1class <- shape_data_handaxe$Class[1:18]
handaxe1pcs <- cbind(handaxe1pcs, handaxe1class)
handaxe1pcs <- rename(handaxe1pcs, Class = handaxe1class)
handaxe1lda <- lda(Class ~ ., handaxe1pcs)
handaxe1ldapredict <- predict(handaxe1lda)
handaxe1ldaplot <- cbind(handaxe1ldapredict$x[, 1:3], handaxe1pcs)
handaxe1ldaplotggplot <- ggplot(handaxe1ldaplot, aes(LD1, LD2)) +
  geom_point(aes(shape = Class), size = 2) +
  labs(x = "LDA 1", y = "LDA 2") + 
  scale_shape_manual(values=c(17,15,3,16)) +
  theme_minimal() + 
  theme(axis.text.x = element_blank(), 
        axis.text.y = element_blank(), 
        legend.position = "none",  
        axis.title.y = element_text(size = 6),  
        axis.title.x = element_text(size = 6))

handaxe2 <- gpa_handaxe$coords[, , 19:36]
df_handaxe2 <- geomorph.data.frame(shape = handaxe2, class = shape_data_handaxe$Class[19:36], artefact = shape_data_handaxe$Artefact[19:36])
H2 <- procD.lm(shape ~ class, data = df_handaxe2, print.progress = FALSE)
summary(H2)

handaxe2pcs <- as.data.frame(pca_handaxe$x[19:36, 1:10]) 
handaxe2class <- shape_data_handaxe$Class[19:36]
handaxe2pcs <- cbind(handaxe2pcs, handaxe2class)
handaxe2pcs <- rename(handaxe2pcs, Class = handaxe2class)
handaxe2lda <- lda(Class ~ ., handaxe2pcs)
handaxe2ldapredict <- predict(handaxe2lda)
handaxe2ldaplot <- cbind(handaxe2ldapredict$x[, 1:3], handaxe2pcs)
handaxe2ldaplotggplot <- ggplot(handaxe2ldaplot, aes(LD1, LD2)) +
  geom_point(aes(shape = Class), size = 2) +
  labs(x = "LDA 1", y = "LDA 2") +
  scale_shape_manual(values=c(17,15,3,16)) +
  theme_minimal() +
  theme(axis.text.x = element_blank(), 
        axis.text.y = element_blank(), 
        legend.position = "none",  
        axis.title.y = element_text(size = 6),  
        axis.title.x = element_text(size = 6))

handaxe3 <- gpa_handaxe$coords[, , 37:54]
df_handaxe3 <- geomorph.data.frame(shape = handaxe3, class = shape_data_handaxe$Class[37:54], artefact = shape_data_handaxe$Artefact[37:54])
H3 <- procD.lm(shape ~ class, data = df_handaxe3, print.progress = FALSE)
summary(H3)

handaxe3pcs <- as.data.frame(pca_handaxe$x[37:54, 1:10])
handaxe3class <- shape_data_handaxe$Class[37:54]
handaxe3pcs <- cbind(handaxe3pcs, handaxe3class)
handaxe3pcs <- rename(handaxe3pcs, Class = handaxe3class)
handaxe3lda <- lda(Class ~ ., handaxe3pcs)
handaxe3ldapredict <- predict(handaxe3lda)
handaxe3ldaplot <- cbind(handaxe3ldapredict$x[, 1:3], handaxe3pcs)
handaxe3ldaplotggplot <- ggplot(handaxe3ldaplot, aes(LD1, LD2)) +
  geom_point(aes(shape = Class), size = 2) +
  labs(x = "LDA 1", y = "LDA 2") + 
  scale_shape_manual(values=c(17,15,3,16)) +
  theme_minimal() +
  theme(axis.text.x = element_blank(), 
        axis.text.y = element_blank(), 
        legend.position = "none",  
        axis.title.y = element_text(size = 6),  
        axis.title.x = element_text(size = 6))

handaxe4 <- gpa_handaxe$coords[, , 55:72] 
df_handaxe4 <- geomorph.data.frame(shape = handaxe4, class = shape_data_handaxe$Class[55:72], artefact = shape_data_handaxe$Artefact[55:72]) 
H4 <- procD.lm(shape ~ class, data = df_handaxe4, print.progress = FALSE) 
summary(H4)

handaxe4pcs <- as.data.frame(pca_handaxe$x[55:72, 1:10])
handaxe4class <- shape_data_handaxe$Class[55:72]
handaxe4pcs <- cbind(handaxe4pcs, handaxe4class)
handaxe4pcs <- rename(handaxe4pcs, Class = handaxe4class)
handaxe4lda <- lda(Class ~ ., handaxe4pcs)
handaxe4ldapredict <- predict(handaxe4lda)
handaxe4ldaplot <- cbind(handaxe4ldapredict$x[, 1:3], handaxe4pcs)
handaxe4ldaplotggplot <- ggplot(handaxe4ldaplot, aes(LD1, LD2)) +
  geom_point(aes(shape = Class), size = 2) +
  labs(x = "LDA 1", y = "LDA 2") +
  scale_shape_manual(values=c(17,15,3,16)) +
  theme_minimal() +
  theme(axis.text.x = element_blank(), 
        axis.text.y = element_blank(), 
        legend.position = "none",  
        axis.title.y = element_text(size = 6),  
        axis.title.x = element_text(size = 6))

handaxe5 <- gpa_handaxe$coords[, , 73:90]
df_handaxe5 <- geomorph.data.frame(shape = handaxe5, class = shape_data_handaxe$Class[73:90], artefact = shape_data_handaxe$Artefact[73:90])
H5 <- procD.lm(shape ~ class, data = df_handaxe5, print.progress = FALSE)
summary(H5)

handaxe5pcs <- as.data.frame(pca_handaxe$x[73:90, 1:10])
handaxe5class <- shape_data_handaxe$Class[73:90]
handaxe5pcs <- cbind(handaxe5pcs, handaxe5class)
handaxe5pcs <- rename(handaxe5pcs, Class = handaxe5class)
handaxe5lda <- lda(Class ~ ., handaxe5pcs)
handaxe5ldapredict <- predict(handaxe5lda)
handaxe5ldaplot <- cbind(handaxe5ldapredict$x[, 1:3], handaxe5pcs)
handaxe5ldaplotggplot <- ggplot(handaxe5ldaplot, aes(LD1, LD2)) +
  geom_point(aes(shape = Class), size = 2) +
  labs(x = "LDA 1", y = "LDA 2") +
  scale_shape_manual(values=c(17,15,3,16)) +
  theme_minimal() + 
  theme(axis.text.x = element_blank(), 
        axis.text.y = element_blank(), 
        legend.position = "none",  
        axis.title.y = element_text(size = 6),  
        axis.title.x = element_text(size = 6))

lda.figure <-  plot_grid(elongated1ldaplotggplot,
                         elongated2ldaplotggplot,
                         elongated3ldaplotggplot, 
                         elongated4ldaplotggplot, 
                         elongated5ldaplotggplot,
                         tanged1ldaplotggplot, 
                         tanged2ldaplotggplot, 
                         tanged3ldaplotggplot, 
                         tanged4ldaplotggplot, 
                         tanged5ldaplotggplot,
                         handaxe1ldaplotggplot,
                         handaxe2ldaplotggplot,
                         handaxe3ldaplotggplot,
                         handaxe4ldaplotggplot,
                         handaxe5ldaplotggplot,
                         ncol = 5, align = 'v')

lda.figure.legend <- get_legend(elongated1ldaplotggplot +
                                  guides(color = guide_legend(nrow = 1)) +
                                  theme(legend.position = "bottom"))

plot_grid(lda.figure, lda.figure.legend, ncol = 1, rel_heights = c(1, .1))

```
  
  
### **Stage 6: Measurement Data (Exploratory Framework)**

To first examine the metric data, bivariate plots for all three artefact groups were generated in ggplot:

```{r, echo = TRUE, eval = TRUE}
metric_data_elongated <- metric_data[which(metric_data$Type=="Elongated"),]
metric_data_handaxe   <- metric_data[which(metric_data$Type=="Handaxe"),]
metric_data_tanged    <- metric_data[which(metric_data$Type=="Tanged"),]

figure_7a <- ggplot(metric_data_elongated, aes(Length_mm, Width_mm, colour = Artefact, shape = Class)) +
  geom_point() + 
  facet_grid(cols = vars(Artefact), scales = "free", labeller=label_both) + 
  labs(x = "Length (mm)", y = "Width (mm)") + 
  scale_shape_manual(values=c(17,15,3,16)) + 
  theme_minimal() +
  theme(axis.text.x = element_blank(), 
        axis.text.y = element_blank(), 
        legend.position = "none")

figure_7b <- ggplot(metric_data_tanged, aes(Length_mm, Width_mm, colour = Artefact, shape = Class)) + 
  geom_point() + 
  facet_grid(cols = vars(Artefact), scales = "free", labeller=label_both) + 
  labs(x = "Length (mm)", y = "Width (mm)") +  
  scale_shape_manual(values=c(17,15,3,16)) + 
  theme_minimal() + 
  theme(axis.text.x = element_blank(), 
        axis.text.y = element_blank(), 
        legend.position = "none")

figure_7c <- ggplot(metric_data_handaxe, aes(Length_mm, Width_mm, colour = Artefact, shape = Class)) +
  geom_point() + 
  facet_grid(cols = vars(Artefact), scales = "free", labeller=label_both) + 
  labs(x = "Length (mm)", y = "Width (mm)") +  
  scale_shape_manual(values=c(17,15,3,16)) +  
  theme_minimal() + 
  theme(axis.text.x = element_blank(), 
        axis.text.y = element_blank(), 
        legend.position = "none")

figure_7 <- plot_grid(figure_7a, 
                      figure_7b, 
                      figure_7c, 
                      labels= "AUTO", 
                      ncol = 1, 
                      align = 'v') 

plot_grid(figure_7, lda.figure.legend, ncol = 1, rel_heights = c(1, .1))

```
  
### **Stage 7: Measurement Data (Analytical Framework)**

```{r, echo = TRUE, eval = TRUE}

metric_data_elongated <- arrange(metric_data_elongated, Artefact, Class)

metric_data_elongated[1:18,]  %>% 
  group_by(Class) %>% 
  summarise(count = n(), 
            mean = mean(Length_mm, na.rm = TRUE), 
            sd = sd(Length_mm, na.rm = TRUE), 
            min = min(Length_mm, na.rm = TRUE), 
            max = max(Length_mm, na.rm = TRUE), 
            cv = sd(Length_mm, na.rm = TRUE)/mean(Length_mm, na.rm = TRUE)*100) %>% 
  dplyr::mutate_if(is.numeric, format, 1)

metric_data_elongated[1:18,]  %>% 
  filter(Class=="Novice" | Class == "Intermediate" | Class == "Professional") %>% 
  summarise(count = n(), 
            cv = sd(Length_mm, na.rm = TRUE)/mean(Length_mm, na.rm = TRUE)*100) %>% 
  dplyr::mutate_if(is.numeric, format, 1)

metric_data_elongated[19:36,] %>% 
  group_by(Class) %>% 
  summarise(count = n(), mean = mean(Length_mm, na.rm = TRUE), 
            sd = sd(Length_mm, na.rm = TRUE), 
            min = min(Length_mm, na.rm = TRUE), 
            max = max(Length_mm, na.rm = TRUE), 
            cv = sd(Length_mm, na.rm = TRUE)/mean(Length_mm, na.rm = TRUE)*100) %>%
  dplyr::mutate_if(is.numeric, format, 1)

metric_data_elongated[19:36,]  %>% 
  filter(Class=="Novice" | Class == "Intermediate" | Class == "Professional") %>%
  summarise(count = n(), 
            cv = sd(Length_mm, na.rm = TRUE)/mean(Length_mm, na.rm = TRUE)*100) %>% 
  dplyr::mutate_if(is.numeric, format, 1)

metric_data_elongated[37:54,] %>% 
  group_by(Class) %>% 
  summarise(count = n(), 
            mean = mean(Length_mm, na.rm = TRUE), 
            sd = sd(Length_mm, na.rm = TRUE), 
            min = min(Length_mm, na.rm = TRUE), 
            max = max(Length_mm, na.rm = TRUE), 
            cv = sd(Length_mm, na.rm = TRUE)/mean(Length_mm, na.rm = TRUE)*100) %>%
  dplyr::mutate_if(is.numeric, format, 1)

metric_data_elongated[37:54,]  %>% 
  filter(Class=="Novice" | Class == "Intermediate" | Class == "Professional") %>% 
  summarise(count = n(), 
            cv = sd(Length_mm, na.rm = TRUE)/mean(Length_mm, na.rm = TRUE)*100) %>% 
  dplyr::mutate_if(is.numeric, format, 1)

metric_data_elongated[55:72,] %>% 
  group_by(Class) %>% 
  summarise(count = n(), 
            mean = mean(Length_mm, na.rm = TRUE), 
            sd = sd(Length_mm, na.rm = TRUE), 
            min = min(Length_mm, na.rm = TRUE), 
            max = max(Length_mm, na.rm = TRUE), 
            cv = sd(Length_mm, na.rm = TRUE)/mean(Length_mm, na.rm = TRUE)*100) %>% 
  dplyr::mutate_if(is.numeric, format, 1)

metric_data_elongated[55:72,]  %>% 
  filter(Class=="Novice" | Class == "Intermediate" | Class == "Professional") %>% 
  summarise(count = n(), 
            cv = sd(Length_mm, na.rm = TRUE)/mean(Length_mm, na.rm = TRUE)*100) %>% 
  dplyr::mutate_if(is.numeric, format, 1)

metric_data_elongated[73:90,] %>% 
  group_by(Class) %>% 
  summarise(count = n(), 
            mean = mean(Length_mm, na.rm = TRUE), 
            sd = sd(Length_mm, na.rm = TRUE), 
            min = min(Length_mm, na.rm = TRUE), 
            max = max(Length_mm, na.rm = TRUE), 
            cv = sd(Length_mm, na.rm = TRUE)/mean(Length_mm, na.rm = TRUE)*100) %>% 
  dplyr::mutate_if(is.numeric, format, 1)

metric_data_elongated[73:90,]  %>% 
  filter(Class=="Novice" | Class == "Intermediate" | Class == "Professional") %>% 
  summarise(count = n(), 
            cv = sd(Length_mm, na.rm = TRUE)/mean(Length_mm, na.rm = TRUE)*100) %>%
  dplyr::mutate_if(is.numeric, format, 1)

shapiro.test(metric_data_elongated$Length_mm[2:4])
shapiro.test(metric_data_elongated$Width_mm[2:4])
shapiro.test(metric_data_elongated$Length_mm[5:17]) 
shapiro.test(metric_data_elongated$Width_mm[5:17])
shapiro.test(metric_data_elongated$Length_mm[20:22])
shapiro.test(metric_data_elongated$Width_mm[20:22]) 
shapiro.test(metric_data_elongated$Length_mm[23:35])
shapiro.test(metric_data_elongated$Width_mm[23:35])
shapiro.test(metric_data_elongated$Length_mm[38:40]) 
shapiro.test(metric_data_elongated$Width_mm[38:40]) 
shapiro.test(metric_data_elongated$Length_mm[41:53])
shapiro.test(metric_data_elongated$Width_mm[41:53])
shapiro.test(metric_data_elongated$Length_mm[56:58])
shapiro.test(metric_data_elongated$Width_mm[56:58])
shapiro.test(metric_data_elongated$Length_mm[59:71])
shapiro.test(metric_data_elongated$Width_mm[59:71])
shapiro.test(metric_data_elongated$Length_mm[74:76])
shapiro.test(metric_data_elongated$Width_mm[74:76])
shapiro.test(metric_data_elongated$Length_mm[77:89])
shapiro.test(metric_data_elongated$Width_mm[77:89])

summary(manova(cbind(Length_mm, Width_mm) ~ Class, 
               data = metric_data_elongated[1:18,]))
summary.aov(manova(cbind(Length_mm, Width_mm) ~ Class, 
                   data = metric_data_elongated[1:18,]))
summary(manova(cbind(Length_mm, Width_mm) ~ Class, 
               data = metric_data_elongated[19:36,]))
summary.aov(manova(cbind(Length_mm, Width_mm) ~ Class, 
                   data = metric_data_elongated[19:36,]))
summary(manova(cbind(Length_mm, Width_mm) ~ Class, 
               data = metric_data_elongated[37:54,]))
summary.aov(manova(cbind(Length_mm, Width_mm) ~ Class, 
                   data = metric_data_elongated[37:54,]))
summary(manova(cbind(Length_mm, Width_mm) ~ Class, 
               data = metric_data_elongated[55:72,]))
summary.aov(manova(cbind(Length_mm, Width_mm) ~ Class, 
                   data = metric_data_elongated[55:72,]))
summary(manova(cbind(Length_mm, Width_mm) ~ Class, 
               data = metric_data_elongated[73:90,]))
summary.aov(manova(cbind(Length_mm, Width_mm) ~ Class, 
                   data = metric_data_elongated[73:90,]))

metric_data_handaxe <- arrange(metric_data_handaxe, Artefact, Class)

metric_data_handaxe[1:18,]  %>% 
  group_by(Class) %>%
  summarise(count = n(), 
            mean = mean(Length_mm, na.rm = TRUE), 
            sd = sd(Length_mm, na.rm = TRUE), 
            min = min(Length_mm, na.rm = TRUE), 
            max = max(Length_mm, na.rm = TRUE), 
            cv = sd(Length_mm, na.rm = TRUE)/mean(Length_mm, na.rm = TRUE)*100) %>% 
  dplyr::mutate_if(is.numeric, format, 1)

metric_data_handaxe[1:18,]  %>% 
  filter(Class=="Novice" | Class == "Intermediate" | Class == "Professional") %>% 
  summarise(count = n(), 
            cv = sd(Length_mm, na.rm = TRUE)/mean(Length_mm, na.rm = TRUE)*100) %>%
  dplyr::mutate_if(is.numeric, format, 1)

metric_data_handaxe[19:36,] %>%
  group_by(Class) %>%
  summarise(count = n(), 
            mean = mean(Length_mm, na.rm = TRUE), 
            sd = sd(Length_mm, na.rm = TRUE), 
            min = min(Length_mm, na.rm = TRUE), 
            max = max(Length_mm, na.rm = TRUE), 
            cv = sd(Length_mm, na.rm = TRUE)/mean(Length_mm, na.rm = TRUE)*100) %>%
  dplyr::mutate_if(is.numeric, format, 1)

metric_data_handaxe[19:36,]  %>% 
  filter(Class=="Novice" | Class == "Intermediate" | Class == "Professional") %>%
  summarise(count = n(), 
            cv = sd(Length_mm, na.rm = TRUE)/mean(Length_mm, na.rm = TRUE)*100) %>%
  dplyr::mutate_if(is.numeric, format, 1)

metric_data_handaxe[37:54,] %>% 
  group_by(Class) %>%
  summarise(count = n(), mean = mean(Length_mm, na.rm = TRUE), sd = sd(Length_mm, na.rm = TRUE), min = min(Length_mm, na.rm = TRUE), max = max(Length_mm, na.rm = TRUE), cv = sd(Length_mm, na.rm = TRUE)/mean(Length_mm, na.rm = TRUE)*100) %>%
  dplyr::mutate_if(is.numeric, format, 1)

metric_data_handaxe[37:54,]  %>%
  filter(Class=="Novice" | Class == "Intermediate" | Class == "Professional") %>%
  summarise(count = n(), cv = sd(Length_mm, na.rm = TRUE)/mean(Length_mm, na.rm = TRUE)*100) %>%
  dplyr::mutate_if(is.numeric, format, 1)

metric_data_handaxe[55:72,] %>%
  group_by(Class) %>%
  summarise(count = n(), 
            mean = mean(Length_mm, na.rm = TRUE), 
            sd = sd(Length_mm, na.rm = TRUE), 
            min = min(Length_mm, na.rm = TRUE), 
            max = max(Length_mm, na.rm = TRUE), 
            cv = sd(Length_mm, na.rm = TRUE)/mean(Length_mm, na.rm = TRUE)*100) %>%
  dplyr::mutate_if(is.numeric, format, 1)

metric_data_handaxe[55:72,]  %>%
  filter(Class=="Novice" | Class == "Intermediate" | Class == "Professional") %>%
  summarise(count = n(), 
            cv = sd(Length_mm, na.rm = TRUE)/mean(Length_mm, na.rm = TRUE)*100) %>%
  dplyr::mutate_if(is.numeric, format, 1)

metric_data_handaxe[73:90,] %>%
  group_by(Class) %>%
  summarise(count = n(), 
            mean = mean(Length_mm, na.rm = TRUE),
            sd = sd(Length_mm, na.rm = TRUE), 
            min = min(Length_mm, na.rm = TRUE), 
            max = max(Length_mm, na.rm = TRUE), 
            cv = sd(Length_mm, na.rm = TRUE)/mean(Length_mm, na.rm = TRUE)*100) %>%
  dplyr::mutate_if(is.numeric, format, 1)

metric_data_handaxe[73:90,]  %>%
  filter(Class=="Novice" | Class == "Intermediate" | Class == "Professional") %>%
  summarise(count = n(), 
            cv = sd(Length_mm, na.rm = TRUE)/mean(Length_mm, na.rm = TRUE)*100) %>%
  dplyr::mutate_if(is.numeric, format, 1)

shapiro.test(metric_data_handaxe$Length_mm[2:4])
shapiro.test(metric_data_handaxe$Width_mm[2:4])
shapiro.test(metric_data_handaxe$Length_mm[5:17])
shapiro.test(metric_data_handaxe$Width_mm[5:17])
shapiro.test(metric_data_handaxe$Length_mm[20:22])
shapiro.test(metric_data_handaxe$Width_mm[20:22])
shapiro.test(metric_data_handaxe$Length_mm[23:35])
shapiro.test(metric_data_handaxe$Width_mm[23:35])
shapiro.test(metric_data_handaxe$Length_mm[38:40])
shapiro.test(metric_data_handaxe$Width_mm[38:40])
shapiro.test(metric_data_handaxe$Length_mm[41:53])
shapiro.test(metric_data_handaxe$Width_mm[41:53])
shapiro.test(metric_data_handaxe$Length_mm[56:58])
shapiro.test(metric_data_handaxe$Width_mm[56:58])
shapiro.test(metric_data_handaxe$Length_mm[59:71])
shapiro.test(metric_data_handaxe$Width_mm[59:71]) 
shapiro.test(metric_data_handaxe$Length_mm[74:76])
shapiro.test(metric_data_handaxe$Width_mm[74:76])
shapiro.test(metric_data_handaxe$Length_mm[77:89])
shapiro.test(metric_data_handaxe$Width_mm[77:89])

summary(manova(cbind(Length_mm, Width_mm) ~ Class, 
               data = metric_data_handaxe[1:18,]))
summary.aov(manova(cbind(Length_mm, Width_mm) ~ Class, 
                   data = metric_data_handaxe[1:18,]))
summary(manova(cbind(Length_mm, Width_mm) ~ Class, 
               data = metric_data_handaxe[19:36,]))
summary.aov(manova(cbind(Length_mm, Width_mm) ~ Class, 
                   data = metric_data_handaxe[19:36,]))
adonis(cbind(Length_mm, Width_mm) ~ Class, 
       data = metric_data_handaxe[37:54,])
summary(manova(cbind(Length_mm, Width_mm) ~ Class, 
               data = metric_data_handaxe[55:72,]))
summary.aov(manova(cbind(Length_mm, Width_mm) ~ Class, 
                   data = metric_data_handaxe[55:72,]))
summary(manova(cbind(Length_mm, Width_mm) ~ Class, 
               data = metric_data_handaxe[73:90,]))
summary.aov(manova(cbind(Length_mm, Width_mm) ~ Class, 
                   data = metric_data_handaxe[73:90,]))

metric_data_tanged <- arrange(metric_data_tanged, Artefact, Class)

metric_data_tanged[1:18,]  %>% 
  group_by(Class) %>%
  summarise(count = n(), 
            mean = mean(Length_mm, na.rm = TRUE), 
            sd = sd(Length_mm, na.rm = TRUE), 
            min = min(Length_mm, na.rm = TRUE), 
            max = max(Length_mm, na.rm = TRUE), 
            cv = sd(Length_mm, na.rm = TRUE)/mean(Length_mm, na.rm = TRUE)*100) %>%
  dplyr::mutate_if(is.numeric, format, 1)

metric_data_tanged[1:18,]  %>%
  filter(Class=="Novice" | Class == "Intermediate" | Class == "Professional") %>%
  summarise(count = n(), 
            cv = sd(Length_mm, na.rm = TRUE)/mean(Length_mm, na.rm = TRUE)*100) %>%
  dplyr::mutate_if(is.numeric, format, 1)

metric_data_tanged[19:36,] %>%
  group_by(Class) %>%
  summarise(count = n(), 
            mean = mean(Length_mm, na.rm = TRUE), 
            sd = sd(Length_mm, na.rm = TRUE), 
            min = min(Length_mm, na.rm = TRUE), 
            max = max(Length_mm, na.rm = TRUE), 
            cv = sd(Length_mm, na.rm = TRUE)/mean(Length_mm, na.rm = TRUE)*100) %>%
  dplyr::mutate_if(is.numeric, format, 1)

metric_data_tanged[19:36,]  %>%
  filter(Class=="Novice" | Class == "Intermediate" | Class == "Professional") %>%
  summarise(count = n(), cv = sd(Length_mm, na.rm = TRUE)/mean(Length_mm, na.rm = TRUE)*100) %>%
  dplyr::mutate_if(is.numeric, format, 1)

metric_data_tanged[37:54,] %>%
  group_by(Class) %>%
  summarise(count = n(), 
            mean = mean(Length_mm, na.rm = TRUE), 
            sd = sd(Length_mm, na.rm = TRUE), 
            min = min(Length_mm, na.rm = TRUE), 
            max = max(Length_mm, na.rm = TRUE), 
            cv = sd(Length_mm, na.rm = TRUE)/mean(Length_mm, na.rm = TRUE)*100) %>%
  dplyr::mutate_if(is.numeric, format, 1)

metric_data_tanged[37:54,]  %>%
  filter(Class=="Novice" | Class == "Intermediate" | Class == "Professional") %>%
  summarise(count = n(), 
            cv = sd(Length_mm, na.rm = TRUE)/mean(Length_mm, na.rm = TRUE)*100) %>%
  dplyr::mutate_if(is.numeric, format, 1)

metric_data_tanged[55:72,] %>%
  group_by(Class) %>%
  summarise(count = n(), 
            mean = mean(Length_mm, na.rm = TRUE), 
            sd = sd(Length_mm, na.rm = TRUE), 
            min = min(Length_mm, na.rm = TRUE), 
            max = max(Length_mm, na.rm = TRUE), 
            cv = sd(Length_mm, na.rm = TRUE)/mean(Length_mm, na.rm = TRUE)*100) %>%
  dplyr::mutate_if(is.numeric, format, 1)

metric_data_tanged[55:72,]  %>%
  filter(Class=="Novice" | Class == "Intermediate" | Class == "Professional") %>%
  summarise(count = n(), 
            cv = sd(Length_mm, na.rm = TRUE)/mean(Length_mm, na.rm = TRUE)*100) %>%
  dplyr::mutate_if(is.numeric, format, 1)

metric_data_tanged[73:90,] %>%
  group_by(Class) %>%
  summarise(count = n(), 
            mean = mean(Length_mm, na.rm = TRUE), 
            sd = sd(Length_mm, na.rm = TRUE), 
            min = min(Length_mm, na.rm = TRUE), 
            max = max(Length_mm, na.rm = TRUE), 
            cv = sd(Length_mm, na.rm = TRUE)/mean(Length_mm, na.rm = TRUE)*100) %>%
  dplyr::mutate_if(is.numeric, format, 1)

metric_data_tanged[73:90,]  %>%
  filter(Class=="Novice" | Class == "Intermediate" | Class == "Professional") %>%
  summarise(count = n(), 
            cv = sd(Length_mm, na.rm = TRUE)/mean(Length_mm, na.rm = TRUE)*100) %>%
  dplyr::mutate_if(is.numeric, format, 1)

shapiro.test(metric_data_tanged$Length_mm[2:4]) 
shapiro.test(metric_data_tanged$Width_mm[2:4])
shapiro.test(metric_data_tanged$Length_mm[5:17])
shapiro.test(metric_data_tanged$Width_mm[5:17])
shapiro.test(metric_data_tanged$Length_mm[20:22])
shapiro.test(metric_data_tanged$Width_mm[20:22])
shapiro.test(metric_data_tanged$Length_mm[23:35])
shapiro.test(metric_data_tanged$Width_mm[23:35])
shapiro.test(metric_data_tanged$Length_mm[38:40])
shapiro.test(metric_data_tanged$Width_mm[38:40])
shapiro.test(metric_data_tanged$Length_mm[41:53])
shapiro.test(metric_data_tanged$Width_mm[41:53])
shapiro.test(metric_data_tanged$Length_mm[56:58])
shapiro.test(metric_data_tanged$Width_mm[56:58])
shapiro.test(metric_data_tanged$Length_mm[59:71])
shapiro.test(metric_data_tanged$Width_mm[59:71])
shapiro.test(metric_data_tanged$Length_mm[74:76]) 
shapiro.test(metric_data_tanged$Width_mm[74:76]) 
shapiro.test(metric_data_tanged$Length_mm[77:89]) 
shapiro.test(metric_data_tanged$Width_mm[77:89]) 

summary(manova(cbind(Length_mm, Width_mm) ~ Class, 
               data = metric_data_tanged[1:18,]))
summary.aov(manova(cbind(Length_mm, Width_mm) ~ Class, 
                   data = metric_data_tanged[1:18,])) 
summary(manova(cbind(Length_mm, Width_mm) ~ Class, 
               data = metric_data_tanged[19:36,]))
summary.aov(manova(cbind(Length_mm, Width_mm) ~ Class, 
                   data = metric_data_tanged[19:36,]))
summary(manova(cbind(Length_mm, Width_mm) ~ Class, 
               data = metric_data_tanged[37:54,]))
summary.aov(manova(cbind(Length_mm, Width_mm) ~ Class, 
                   data = metric_data_tanged[37:54,]))
summary(manova(cbind(Length_mm, Width_mm) ~ Class, 
               data = metric_data_tanged[55:72,]))
summary.aov(manova(cbind(Length_mm, Width_mm) ~ Class, 
                   data = metric_data_tanged[55:72,]))
summary(manova(cbind(Length_mm, Width_mm) ~ Class, 
               data = metric_data_tanged[73:90,]))
summary.aov(manova(cbind(Length_mm, Width_mm) ~ Class, 
                   data = metric_data_tanged[73:90,]))
```

### **Concluding Remarks**

For any queries, or for more information on any aspect of this markdown, please contact C.S.Hoggard@soton.ac.uk. 

### **Acknowledgements**  

We thank Moesgaard Museum for loaning the artefacts used throughout this study. We would also like to thank all illustrators for their time during this experiment. CSH and FR thank the Independent Research Fund Denmark for grant #610700059B and FR also gratefully acknowledges funding the European Research Council (grant agreement 817564 under the Horizon 2020 research and innovation programme). We also thank John McNabb for comments on earlier drafts of this paper.